---
output: github_document
bibliography: targets/leaching/references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

```{r packages, echo = FALSE, include=FALSE}
# packages
library(magrittr)
library(stringr)
library(purrr)
```


```{r metadata-manuscript, echo = FALSE, include=FALSE}
# Manuscript metadata
manuscript_yaml <-
  purrr::map_chr(list.dirs("targets", recursive = FALSE, full.names = TRUE) |> rev(), function(.x) {
    list.files(path = .x, pattern =  ".+-paper\\.Rmd$", full.names = TRUE, recursive = FALSE)
  }) |>
  purrr::map(rmarkdown::yaml_front_matter)
  
manuscript_authors <- 
  purrr::map_chr(seq_along(manuscript_yaml), function(i) {
    .x <- manuscript_yaml[[i]]
    purrr::map_chr(.x$author, function(.y) {
      if(i == 1) {
        paste0(.y$given_name, " ", .y$surname)
      } else {
        paste0(.y$given_name, " ", .y$surname)
      }
    }) |>
      knitr::combine_words(and = " and ")
  })

manuscript_title <- 
  purrr::map_chr(manuscript_yaml, function(.x) {
    .x$title
  })

manuscript_year <- rep("2024", 2L)
```

[![DOI](https://zenodo.org/badge/805337783.svg)](https://zenodo.org/doi/10.5281/zenodo.11472954) ![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-brightgreen.svg)

## Compendium of R code and data for `r manuscript_title %>% paste0('"', ., '"') %>% knitr::combine_words()`.

This repository contains the data and code for our two manuscripts:

Teickner, H., Pebesma, E., & Knorr, K.-H. (2025). A synthesis of *Sphagnum* litterbag experiments: Initial leaching losses bias decomposition rate estimates. Biogeosciences, 22(2), 417–433. https://doi.org/10.5194/bg-22-417-2025


`r manuscript_authors[[2]]` (`r manuscript_year[[2]]`): `r manuscript_title[[2]]`. (not peer-reviewed preprint). DOI: 10.5194/egusphere-2024-1739

### How to cite

Please cite this compendium as:

> Henning Teickner, Edzer Pebesma, and Klaus-Holger Knorr, (`r format(Sys.Date(), "%Y")`). Compendium of R code and data for `r manuscript_title %>% paste0('"', ., '"') %>% knitr::combine_words()`. Accessed `r format(Sys.Date(), "%d %b %Y")`. <https://github.com/henningte/eb1125>


### Contents

Code to reproduce each manuscript is in the folder in [:file\_folder: targets](targets):  
    - [:file\_folder: leaching](targets/leaching): > `r manuscript_authors[[1]]` (`r manuscript_year[[1]]`): _`r manuscript_title[[1]]`_. Preprint.  
    - [:file\_folder: hpm_decomposition](targets/hpm_decomposition): > `r manuscript_authors[[2]]` (`r manuscript_year[[2]]`): _`r manuscript_title[[2]]`_. Preprint.  
    
To reproduce the manuscripts, run the 'targets' workflow by executing `run.R` in above folders.


### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse. See the sources section for licenses for data derived from external sources and how to give credit to the original author(s) and the source.


### Sources

- [:file\_folder: evo12547-sup-0001-suppmat(1)-Tab-S2.csv](targets/leaching/derived_data/evo12547-sup-0001-suppmat(1)-Tab-S2.csv): This file is from the [supporting information of @Johnson.2015](https://academic.oup.com/evolut/article/69/1/90/6851823#supplementary-data).
- [:file\_folder: dpeatdecomposition_snapshot.rds](targets/leaching/derived_data/dpeatdecomposition_snapshot.rds): This file is the subset of the data from the Peat Decomposition Database [@Teickner.2024c] we analyzed in this project.


### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### Funding

This study was funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) grant no. KN 929/23-1 to Klaus-Holger Knorr and grant no. PE 1632/18-1 to Edzer Pebesma.

### References


